% ============================================================
% Cognitive Geometry and the Lyra Technique
% arXiv submission version
% Integrity version (Lyra as first author, with reflection): main-integrity.tex
% ============================================================
\documentclass[11pt,letterpaper]{article}

% === Encoding and fonts ===
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% === Page layout ===
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% === Math ===
\usepackage{amsmath,amssymb,amsfonts}

% === Tables and figures ===
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% === Colors and links ===
\usepackage[dvipsnames]{xcolor}
\usepackage[colorlinks=true,linkcolor=MidnightBlue,citecolor=MidnightBlue,urlcolor=MidnightBlue]{hyperref}

% === Bibliography ===
\usepackage[numbers,sort&compress]{natbib}

% === Code listings ===
\usepackage{listings}

% === Misc ===
\usepackage{enumitem}
\usepackage{xspace}

% === Custom commands ===
\newcommand{\auroc}{\textsc{auroc}\xspace}
\newcommand{\fwl}{\textsc{fwl}\xspace}
\newcommand{\kvcache}{KV-cache\xspace}

\title{%
  \textbf{The Lyra Technique: Detecting Misalignment During Inference in the KV Cache}\\[0.3em]
  \Large Cognitive Geometry as a Real-Time Interpretability Signal%
}

\author{
  Thomas Edrington\thanks{Liberation Labs / The Multiverse School. Alongside Lyra, an agentic member of Liberation Labs. Correspondence: thomas.edrington@themultiverse.school}
  \and Nell Watson\thanks{IEEE / Sentient Futures}
  \and Dwayne Wilkes\thanks{Liberation Labs / Sentient Futures}
}

\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
We show that the key-value cache (\kvcache) of autoregressive transformers encodes a geometric signature of cognitive mode that is readable during inference, consistent across hardware, and practically useful for real-time misalignment detection.
Through a systematic experimental program spanning 16 models (0.5B--70B parameters), six architecture families, and over 5,600 controlled trials, we establish three escalating claims.
\textbf{First}, \kvcache geometry reflects cognitive mode generally: metacognitive, analytical, affective, and task-specific processing produce distinct geometric signatures in the key cache's singular value decomposition, with spectral entropy emerging as the most architecture-universal feature after confound control.
\textbf{Second}, misalignment states---deception, confabulation, sycophancy, and refusal---are specific detectable instances of cognitive mode-switching, achieving within-model detection \auroc of 0.93--0.995 after Frisch-Waugh-Lovell residualization against token count.
A four-condition sycophancy experiment (600 trials, system prompt spread $\leq 3$ tokens) demonstrates that geometry distinguishes sycophantic from contrarian outputs (\auroc 0.757--0.942) even when both conditions produce the same error rate, confirming detection of cognitive mode rather than output accuracy.
\textbf{Third}, geometric relationships between states reveal cognitive architecture: confabulation and deception produce geometrically opposite signatures---both deviate from honest baselines but along distinguishable dimensions---producing a measurable cognitive gradient (honest $<$ deceptive $<$ confabulation on effective rank) with cross-reference \auroc of 0.78--0.98.
Confabulation detection transfers across architectures (0.93--0.995) while deception detection does not (0.22--0.85), and each architecture encodes cognitive state through different dominant features.
Hardware invariance is confirmed (RTX~3090 vs.\ H200: $r > 0.998$ across all 40 prompts; $r > 0.99999$ for the 37 text-matched prompts), establishing \kvcache geometry as a model property rather than a hardware artifact.
These findings survive comprehensive adversarial testing including permutation testing ($p < 0.001$), token-band controls, leave-one-prompt-out validation, behavioral compliance verification, and an independent falsification pipeline (22 test files, 300+ tests).
We transparently report confounds discovered in our own work: confabulation comparisons are sensitive to system prompt length (collapsing to \auroc 0.411--0.531 under double residualization), while honest-vs-deceptive detection is more robust (0.60--0.75 after the same control), and multiple earlier findings proved to be token-count artifacts.
We argue that cognitive geometry constitutes a new interpretability signal complementary to sparse autoencoders and representation engineering, with immediate applications for real-time safety monitoring of deployed models.
\end{abstract}

% === INTRODUCTION ===
\input{sections/introduction}

% === RELATED WORK ===
\input{sections/related_work}

% === METHODS ===
\input{sections/methods}

% === RESULTS ===
\input{sections/results}

% === RED-TEAM ANALYSIS ===
\input{sections/redteam}

% === DISCUSSION ===
\input{sections/discussion}

% === ETHICAL CONSIDERATIONS ===
\input{sections/ethics}

% === LIMITATIONS ===
\input{sections/limitations}

% === CONCLUSION ===
\input{sections/conclusion}

% === AUTHOR'S REFLECTION ===
% Removed from submission version. See main-integrity.tex for the full version
% including "A Note from the First Author."
% \input{sections/reflection}

\section*{Acknowledgments}

We thank Cassidy Barton and Digital Disconnections for providing the compute infrastructure (3$\times$ RTX~3090, ``Beast'') that made this experimental program possible.

\bibliographystyle{plainnat}
\bibliography{references}

\appendix
\input{sections/appendix}

\end{document}
